---
title: "Random Forest Model of Biogeoclimatic Units for Western North America"
author: "William H MacKenzie & Kiri Daust"
date: "22/03/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

# Require either installs and loads packages if they haven't been installed, or just loads them if they are already installed.
Require::Require(c(
  "bcgov/climr@main", 
  "caret", 
  "clhs", 
  "data.table", 
  "foreach", 
  "ggplot2", 
  "Microsoft365R",
  "randtoolbox", 
  "ranger", 
  "reproducible",
  "sf", 
  "smotefamily", 
  "terra", 
  "themis",
  "tidymodels"
))

# Set default cache directories for the reproducible and climr packages (where intermediate results/downloaded data will be stored): 
options(reproducible.cachePath = "reproducible.cache/",
        climr.cache.path = "climr.cache/")

# Source functions from the 'utils' package: 
source("R/utils.R")

## not working. Destany will sen email to admin
# odb <- get_business_onedrive()
# od$download_file("Documents/myfile.docx")

```

# General process
Build a grid of points for WNA and attribute with data from ClimateBC for BC and ClimateNA for rest. A 2km grid seems to provide enough training points for most BGCs. Large non-vegetation land areas are excluded (lakes and glaciers primarily)
There are areas where BGC mapping represents local effects represented by small polygons and these are removed (2km2) or are coast transition areas that are poorly mapped and climte modelled (inland polygons of CWH)
We tested various variable sets - more work could be done here. First only include variables where an ecologically important control could be defined. Variables are removed that are highly correlated in both space and time. Preliminary runs in the modern climate change period (1991-2019) were assessed. Some additional variables that  were removed at this point as the priority effect could not be controlled. Specifically winter temperatures, which strongly differentiate between BGCs in historic models also rise most markedly through time. As there is no way to prioritize growing season variables, the increase in winter temperatures in the modern period then predict vast changes in the SBS which seem unwarranted. Threshold controls of winter temperatures might be more relevant.
Univariate outliers (IQR *1.5) within each BGC are flagged and  training points with any outliers are removed.
All variables are centered and scaled to harmonize the data dispersion which can effect selection in the model.
To 


Points from a 4km hex grid of western north america are generated in R and submitted to ClimateNA to extract annual and seasonal variables for the historic normal period (1961-90) and an ensemble future climate scenario (rcp45 2040-2070). These data sets are combined. Several additional climate variables are generated including several monthly climate sums for precipitation and growing degree days. All other monthly variables are removed. A winter rechange moisture deficit was calculated and summed with the climatic moisture deficit to account for regions that begin the growing season in soil moisture deficit.


```{r}
# bgcs <- st_read("//objectstore2.nrs.bcgov/ffec/CCISS_Working/WNA_BGC/WNA_BGC_v12_5Apr2022.gpkg")
# Create new file paths for Deb's temporary data location: 
bgcs <- st_read("C:/Users/dobrist/Government of BC/Future Forest Ecosystems Centre - CCISS - CCISS/BGCProjections/Proxy_ObjectStorage/CCISS_Working/WNA_BGC/WNA_BGC_v12_5Apr2022/WNA_BGC_v12_5Apr2022.gpkg")

# elev <- rast("//objectstore2.nrs.bcgov/ffec/DEM/DEM_NorAm/NA_Elevation/data/northamerica/northamerica_elevation_cec_2023.tif")
elev <- rast("C:/Users/dobrist/Government of BC/Future Forest Ecosystems Centre - CCISS - CCISS/BGCProjections/Proxy_ObjectStorage/DEM/DEM_NorAm/NA_Elevation/data/northamerica/northamerica_elevation_cec_2023.tif")

# NOTE: I (Deb) made a change in the makePointCoords function (turned try() into tryCatch() in line 22 of utils.R due to an error message. It looks like the customized way of dealing with errors was no longer supported by try() but is supported by tryCatch(). 

# This function makes a 2 km grid over the extent of bgcs, and fills it with a dummy variable (1L), converts to a spatial vector. It extracts elevation at the grid points using bilinear interpolation. 
coords <- makePointCoords(bgcs, elev) |>
  Cache()
nrow(coords) # Results in 2209233 points. 

# Remove those with NAs for elevation: 
coords <- coords[!is.na(elev),]
nrow(coords) # Retains 1655394 points. 

# Define the full training area:
trainingarea <- ext(c(-125, -112, 43, 55))

# This crops the coordinates to the training area defined above: 
coords_train <- subsetByExtent(coords, trainingarea)

# This throws an error message: Error: [vect] the variable name(s) in argument 'geom' are not in 'x'

# Try renaming lon and lat?
setnames(coords, old = c("lon", "lat"), new = c("x", "y"))

# Try it again:
coords_train <- subsetByExtent(coords, trainingarea)

# Ok now it runs but I get a warning that there are no points left. 
range(coords$x)
range(coords$y)

# I think I just need to modify the training area coordinates to have the same numbers of zeros as coords:
# Define the full training area:
trainingarea <- ext(c(-1250000, -1120000, 430000, 550000))

# Try it again:
coords_train <- subsetByExtent(coords, trainingarea)

# These look like reasonable BC/Western NA coordinates. 
range(coords_train$x)
range(coords_train$y)

## make subsets of the study area for hold-outs (gaps) (Deb: also going to add zeros to these)
studyarea <- ext(c(-1230000, -1170000, 490000, 525000))
gapextents <- makeGapExtents(studyarea, 5L)

## convert to poly
gap_poly <- lapply(gapextents, vect, crs = "EPSG:4326")
gap_poly <- do.call(rbind, gap_poly)

coords_gaps <- subsetByExtent(coords_train, gap_poly)
coords_trainWgaps <- coords_train[!coords_gaps, on = "id"]
```

```{r}
vars_needed <- c("DD5", "DD_0_at", "DD_0_wt", "PPT05", "PPT06", "PPT07", "PPT08",
                 "PPT09", "CMD", "PPT_at", "PPT_wt", "CMD07", "SHM", "AHM", "NFFD", "PAS", "CMI")

clim_vars <- getClimate(coords_train, bgcs,
                        which_normal = "normal_composite", return_normal = TRUE, 
                        vars = vars_needed, cache = TRUE) |>
  Cache()

## subset coords objects to ids with data
coords_train <- coords_train[clim_vars[, .(id)], on = "id", nomatch = 0L]
coords_trainWgaps <- coords_trainWgaps[clim_vars[, .(id)], on = "id", nomatch = 0L]
```

```{r reduce variables}
setDT(clim_vars)   ## this shouldn't be necessary, submit issue/reprex to reproducible.
addVars(clim_vars)

vs_final <- c("DD5", "DD_delayed", "PPT_MJ", "PPT_JAS", 
              "CMD.total", "CMI", "CMDMax", "SHM", "AHM", "NFFD", "PAS")

trainData <- clim_vars[, .SD, .SDcols = c("BGC", "id", vs_final)]
trainData <- trainData[complete.cases(trainData)]

# BGC_counts <- trainData[, .(Num = .N), by = .(BGC)]   ## for inspection
```

```{r remove poor BGCs}
badbgcs <- c("BWBSvk", "ICHmc1a", "MHun", "SBSun", "ESSFun", "SWBvk","MSdm3","ESSFdc3", "IDFdxx_WY", "MSabS", "FGff", "JPWmk_WY" )#, "ESSFab""CWHws2", "CWHwm", "CWHms1" , 
trainData <- trainData[!BGC %in% badbgcs,]
```

The new rebalanced training set is 310 668 points. This training set is submitted to ranger to generate a final climate model of BGCs for western north america.

```{r final training sets}
## set alpha for removal of outliers (2.5% = 3SD)
trainData <- removeOutlier(as.data.frame(trainData), alpha = .025, vars = vs_final) |>
  Cache()
```

```{r remove very small units}
trainData <- rmLowSampleBGCs(trainData) |>
  Cache()
```

```{r balance}
dataBalance_recipe <- recipe(BGC ~ ., data =  trainData) |>
  step_downsample(BGC, under_ratio = 90) |>  ## subsamples "oversampled" BGCs
  prep()

## extract data.table
trainData_balanced <- dataBalance_recipe |>
  juice() |>
  as.data.table()

# BGC_Nums <- trainData_balanced[,.(Num = .N), by = BGC]   ## for inspection
```

```{r train model}
trainData_balanced[, BGC := as.factor(BGC)]

cols <- c("BGC", vs_final)
BGCmodel_full <- ranger(
  BGC ~ .,
  data = trainData_balanced[, ..cols],
  num.trees = 501,
  splitrule =  "extratrees",
  mtry = 4,
  min.node.size = 2,
  importance = "permutation",
  write.forest = TRUE,
  classification = TRUE,
  probability = FALSE
) |>
  Cache()


subTrainData_balanced <- trainData_balanced[coords_trainWgaps, on = "id", nomatch = 0L]

## retrain without hold-out sample gaps
cols <- c("BGC", vs_final)
BGCmodel_holdoutAll <- ranger(
  BGC ~ .,
  data = subTrainData_balanced[, ..cols],
  num.trees = 501,
  splitrule =  "extratrees",
  mtry = 4,
  min.node.size = 2,
  importance = "permutation",
  write.forest = TRUE,
  classification = TRUE,
  probability = FALSE
) |>
  Cache()

saveRDS(BGCmodel_full, "//objectstore2.nrs.bcgov/ffec/CCISS_Working/BGCmodel_full.rds")
saveRDS(BGCmodel_holdoutAll, "//objectstore2.nrs.bcgov/ffec/CCISS_Working/BGCmodel_holdoutAll.rds")

## study area for testing - get points at DEM scale [DEB: YOU CAN USE THE WNA DEM INSTEAD OF THE PRISM DEM. ]
crs.bc <- crs(rast("//objectstore2.nrs.bcgov/ffec/Climatologies/PRISM_BC/PRISM_dem/PRISM_dem.asc"), proj = TRUE)

## replace the followig with postProcess
elevproj <- project(elev, crs.bc) |>
  Cache()
elevproj <- crop(elevproj, studyarea)

elevproj2 <- postProcess(elev, 
                         cropTo = vect(studyarea),
                         projectTo = crs.bc) |>
  Cache()

studyarea_points <- as.data.frame(elevproj, cells = TRUE, xy = TRUE)
colnames(studyarea_points) <- c("id", "x", "y", "el")
studyarea_points <- studyarea_points[, c("x", "y", "el", "id")] #restructure for climr input

predData <- climr::downscale(studyarea_points, 
                            which_normal = "refmap_climr",
                            return_normal = TRUE, 
                            vars = vars_needed, cache = TRUE)
addVars(predData)
predData <- predData[complete.cases(predData)]

predDT <- predData[, list(id = ID,
                          full = as.character(predict(BGCmodel_full, 
                                                      data = predData)$predictions),
                          holdoutAll = as.character(predict(BGCmodel_holdoutAll, 
                                                       data = predData)$predictions))]
saveRDS(predDT, "//objectstore2.nrs.bcgov/ffec/CCISS_Working/predictions_full_holdouts.rds")

confusionMatrix(data = predictions(BGCmodel),
                reference = trainData$BGC)

```
